Measurement and Probability in Fuzzy Quantum Theories

### 2016-11-26 proposal/prop.tex
don-ish

\paragraph*{Research Question.} add discrete theory to the story; do
not assume anything is infinitely precise not even the hypothetical state

Given quantum state as interactive system
and interval-valued probability
and perhaps additionally discrete states in finite fields (try with or without)
investigate gleason (born), bell, KS, and revisit Mermin-Meyer and
other debates and paradoxes

mention aaronson three statements paradox. something is wrong;
investigate; investigate complexity; p=np; mention usat and mention
deutsch-josza improvement depends on precision not as dramatic

\paragraph*{Research Question.}

If power of observer is limited with respect to power of quantum state
then quantum state can always ``fool'' you and get away with things;
you don't have the resources to catch it in an inconsistent state; if
state is interactive and always ``uncommitted'' until you make your
move you can never catch it in a lie

explain secret key exchange; if attacker has limited power cannot
crack it; complexity crucial.

may find surprises like we found with UNIQUE-SAT; not sure where that belongs

clear goals: KS with interval probability or with quantum states as
processes

talk about finite fields; previous work; unique-sat surprise; etc

... and quantum measurement expressed using game semantics

This is really about unifying classical and quantum theories of
measurement. One of the most well-developed and formalized theories of
classical measurement is the one for programming language
observational contextual equivalence.

This worldview can be maintained as long as the impact of the
measurements is really negligible on the execution time of the
functions of interest.

measurement affects runtime and hence affects other measurements;
could even conceive of programs whose input-output behavior is
affected by measurement

This intended level of abstraction is reflected in the first
line of the function declaration in Haskell. This \emph{interface} (or \emph{type}
or \emph{signature} or \emph{contract}) line:
\begin{quote}
\begin{verbatim}
square :: Int -> Int
\end{verbatim}
\end{quote}
summarizes the publicly stated, computationally relevant, aspects of the
function. In this case, it is announced that the function is a computer
realization of a pure mathematical map from one \verb|Int| to another. The
fact that executing the computer realization of the map on a particular input
might consume 10\% of the battery power on your laptop is something
that---although potentially critical knowledge in some circumstances---is not
even expressible in the notation above.

%%-------------------------------------
\subsection*{Partial Functions (Time)}

Computations take time. At some level of abstraction, we can ignore the time
taken by a computation. For example, we can pretend that a calculation that
takes one or two microseconds is similar to the ``instantaneous'' lookup in a
mathematical mapping. But we can hardly keep pretending this as the time used
by the calculation tends towards infinity.

Consider the following Java method:
\begin{quote}
\begin{verbatim}
int collatz (int n) {
  if (n == 1) return 0; 
  else if (n % 2 == 0) return 1 + collatz (n / 2);
  else return 1 + collatz (3 * n + 1);
}
\end{verbatim}
\end{quote}
As evident from the use of \verb|collatz| in the body of the method, this
method is \emph{recursive}. It takes an input number \verb|n| and:
\begin{itemize}
\item if the input number is equal to one, the method immediately stops with
the value zero, which represents in this case the total number of recursive
calls performed before stopping;
\item if the input number is even, the method divides it by two and
recursively invokes itself, adding one to the total number of recursive
calls;
\item if the input number is odd, the method multiplies it by three, adds
one, and then recursively invokes itself, after again adding one to the total
number of recursive calls.
\end{itemize}
It is a famous mathematical conjecture, \emph{the Collatz conjecture}, that
this calculation always terminates for all positive integers. One can quickly
check a few cases:
\begin{itemize}
\item invoking \verb|collatz(3)| iterates through 10, 5, 16, 8, 4, 2, 1, and
  then stops returning the result 7.
\item invoking \verb|collatz(17)| iterates through 52, 26, 13, 40, 20, 10, 5,
16, 8, 4, 2, 1, and then stops returning the result 12.
\end{itemize}

There is however no known proof that this method always stops. Thus, even
though the method has the same interface as the method \verb|square| above,
one may not be able to interpret the two interfaces in the same way. To
account for the possibility that the calculation never stops, the interface:
\begin{quote}
\begin{verbatim}
int collatz (int n)
\end{verbatim}
\end{quote}
should in general be read as saying:
\begin{quote}
Given a particular \verb|int| the method \emph{may or may not} return another
\verb|int| in a finite amount of time.
\end{quote}

Another way to look at the situation is that we have introduced ``time'' as
an implicit parameter of computations. Some computations as the method
\verb|square| update---as a side effect---this time parameter by consuming a
few ticks. By general agreement, this consumption is deemed negligible and
not ``observable'' and does not generally constitute a computational
effect. But when the consumption of time is unbounded, it arguably becomes an
observable computational effect.

%%-------------------------------------
\subsection*{Assignments (Memory)}

Computations require memory. Again, at some level of abstraction we can
pretend that memory does not exist. But what happens if a computation
starts changing external memory locations?

Consider the following variant of the method \verb.square.:
\begin{quote}
\begin{verbatim}
int square (int n) {
  count++;
  return n * n;
}
\end{verbatim}
\end{quote}
As before, the method takes an integer \verb.n. from the domain of Java
integers \verb.int. and returns another integer whose value is the square of
the input. But, as a side effect the method reads and updates an external
reference \verb.count., which can be used to find out how many times the
method was called. Thus if during the execution of a particular program the
method \verb.square. is called ten times, then the value of \verb.count. will
be ten.

Thus the interface:
\begin{quote}
\begin{verbatim}
int square (int n)
\end{verbatim}
\end{quote}
should now be read as saying:
\begin{quote}
Given a particular \verb|int| the method \emph{may possibly} read and update
external memory references and then it \emph{may or may not} return another
\verb|int| in a finite amount of time.
\end{quote}

Another way to look at the situation is that we have introduced ``memory'' as
an implicit parameter of computations. Some computations as the original
method \verb|square| might---as a side effect---allocate, update, and read
several memory locations (registers, cache entries, stack frames, etc.). By
general agreement, these actions on memory are not considered observable and
do not generally constitute a computational effect. But when a computation
updates an external memory location that other program fragments can inspect,
the update of the memory location becomes an observable computational effect.

%%-------------------------------------
\subsection*{Runtime Exceptions (Context)}

Computations occur in a context. When a computation occurs in the context of
a larger computation, it is typically expected to return its result to that
outer computation. But what if a computation cannot return a sensible value
to its context and decides instead to abort it?

Consider again the method \verb|square| whose interface is:
\begin{quote}
\begin{verbatim} 
int square (int n)
\end{verbatim}
\end{quote}
We initially argued that it is possible to interpret this interface to say
that given a particular \verb|int| the method's execution promises to return
another \verb|int|. This is clearly an idealized view: in real life, any of
an unlimited number of failures may occur in the hardware or underlying
software components including the operating system or the Java virtual
machine. The Java language collectively refers to these conditions as
\verb|Error|s or \verb|RuntimeException|s.

So to be more accurate, and putting together our observations so far, one
should interpret the interface:
\begin{quote}
\begin{verbatim} 
int square (int n)
\end{verbatim}
\end{quote}
as saying:
\begin{quote}
Given a particular \verb|int| the method \emph{may possibly} read and
update external references, and then it \emph{may or may not} return another
\verb|int| in a finite time or it \emph{may fail to return to its context}
due to any of a number of errors and runtime exceptions.
\end{quote}

Another way to look at the situation is that we have introduced the
``context'' as an implicit parameter of computations. It is expected that
every inner computation would resume its context with its result. But when an
inner computation encounters a fatal error condition, it can ignore the
context parameter, which causes the error condition to immediately terminate
the entire computation. This escape clearly constitutes an observable
computational effect.

%%-------------------------------------
\subsection*{Other Computational Effects}

The unifying theme in all the above effects is that computations do not
happen in isolation in some abstract mathematical world: rather computations
are realized in some physical computing platform. The physical interactions
of a computation with the computing platform and with other computations may
be ignored in some idealized situation corresponding to simple computations,
but they become crucial if one is interested in modeling the actual behavior
of a sophisticated realistic computation. If these interactions happen
``behind the scene'' they are considered side effects of the computation.

Without further constraints on computations and their contexts, it is
impossible to enumerate all the possible side effects that a computation may
have. The reader may find it instructive to consider the following list of
effects that could be performed by a method with the interface:
\begin{quote}
\begin{verbatim} 
int square (int n)
\end{verbatim}
\end{quote}
Given a particular \verb|int|, the method's behavior may include various
combinations of the following actions:
\begin{itemize}
\item Reading and updating external memory locations, environment parameters,
local or remote file systems or databases, etc.
\item Communicating with a user or another local or remote process.
\item Suspending itself for a certain period of time or until a certain
outside condition happens.
\item Resuming other suspended computations.
\item Loading, compiling, and executing some other computation.
\item Using reflection to examine and perhaps modify its own machine
representation.
\item Producing multiple answers by arranging for one \verb|int| to be
returned now and subsequent answers to be returned when the method is called
again (with the same parameters). 
\item Checkpointing its entire state to a log, which can be examined and
resumed later.
\item Failing due to any of a number of errors and runtime exceptions.
\item Consuming half of the battery life of a laptop and generating enough
heat to keep you warm on a cold winter night.
\end{itemize}
And yes by the way, the method may actually return another \verb|int| in a
finite amount of time.

It is clear that the method's interface is nowhere close to giving us a
reasonable approximation of the possible behaviors of the method: it sweeps
too many things under the rug. Pragmatically, if a programmer invokes a
method with the interface:
\begin{quote}
\begin{verbatim} 
int square (int n)
\end{verbatim}
\end{quote}
the implicit computational effects of the method might come as a
surprise.

The situation is not unique to Java. Almost every other programming
language allows similar computational effects to occur
implicitly. Even hardware circuits and assembly language instructions
often perform implicit effects. For example, an assembly instruction
might change the contents of the flag register as a side effect. In
fact natural languages also include various ``non-compositional
phenomena,'' which are essentially computational effects. For this
reason, programmers are universally taught to avoid side effects if
possible and to be careful when analyzing code that might contain side
effects. But when side effects are unavoidable, the solution is to
\emph{expose them} as the rest of the article explains.

Common wisdom is that classical physics has no randomness. Only
ignorance of the fine details or lack of control over them causes
\emph{statistical randomness}. In principle, but not in practice,
randomness is absent from classical physics.

In sharp contrast, quantum randomness is unavoidable even in principle.

But but but ... Nature is one. The distinction between ``classical''
and ``quantum'' can only be in our minds, or more precisely in our
mathematical models of the universe. We have two mathematical models,
one which is, in principle deterministic, and one which is
probabilistic. Nature must really be one or the other.

Consider this unsettling quote from QBism:

\begin{quote} 
Instead, the wavefunction obeys two fundamentally different laws:
\begin{enumerate}
\item As long as the electron is left to its own devices unobserved
  its wavefunction unfolds smoothly, continuously, and predictably. It
  obeys fixed mathematical laws, just like a bullet in flight and a
  wave on a lake.
\item When the electron reveals its whereabouts by leaving a dot on a
  screen, the wavefunction ``collapses'' suddenly into a new much more
  compact form concentrated on the point of impact.
\end{enumerate}

\end{quote}

Can we develop a mathematical model that unifies both the classical
deterministic perspective and the quantum probabilistic perspective?
Quantum information and quantum computation opens the door to using
computational insights. The main one is that, in a computation,
everything produced during a computation and everything observed about
a computation is finite. The computational perspective imposes a
principle we might call the principle of ``finitely communicable
evidence.''

So imprecision in classical physics becomes, not some second-class
principle that is considered to stem from practical limitations but
rather a foundational principle. If everything is ``fuzzy'' so to
speak and if our mathematical models embrace this fuzziness then a
different worldview emerges. That worldview agrees with classical
physics but embraces imprecision within the mathematical model not as
an external non-mathematical nuisance. This shift in perspective
enables a smooth transition to the world of quantum physics since our
mathematical model has already embraced probabilistic effects.

Three insights:

\begin{itemize}
\item Interactive systems: response of quantum system determined
  \emph{after} the observer initiates the experiment used to observe
  the system
\item Bayesian probability, personal
\item Finite precision
\end{itemize}

Goal: build computational model of QM that encompasses classical
(reversible), computation and observation with finite resources, and
hopefully revolutionize ``quantum thinking''

>>>>

\paragraph*{Research Question.} A long term, ultimate, goal of this
research proposal would be to combine the various pieces into a
coherent computational model of quantum mechanics that keeps explicit
track of resources needed to represent the interactive system
representing the quantum state, its evolution, the observables, and
the probabilities. Such a computational model of quantum mechanics
would encompass classical (reversible) computation and potentially
revolutionize quantum thinking with finite resources. The explicit
manipulation of resources enables one of key idea that emerged in our
previous work: the resources used to represent the quantum system are
in principle unrelated to the resources available to the various
observers. We conjecture that this asymmetry may prove key to
resolving some of the quantum paradoxes.
